### 2018-01-12 ident/docs/ident_manuscript.tex
chnages to results - added new section(s)

\section{Results:}
First in Section \ref{sec:closedform}, we discuss the ability to apply our methods to different nonlinear kinetic rate law formulations within the context of the gluconeogenic model. %Then, in Section we discuss the different experimental design strategies that were found through our identifiability analysis to enable kinetic parameter estimation for the each flux in the model using steady state data. Finally, in Section we expand the application of our method to a relatively large (39 reactions) kinetic model of the red blood cell metabolism.
\subsection{Getting closed-form expressions for each flux in Kotte model:}\label{sec:closedform}
As demonstrated for identifying parameters of $v_1$ earlier in Section \ref{sec:example}, the computational effort required to establish identifiability is governed by the nonlinear complexity of the enzyme kinetic rate law used to model a specific flux. Although computer algebra systems are capable of handling complex symbolic calculations, sometimes, the complexity of getting closed form expressions for all kinetic parameters of certain rate law formulations is too much for the computer algebra system to overcome. We encounter this scenario when expression for the the allosterically activated reaction $v_3$ is described by the MWC model for allosteric regulation. We found that the complexity of the MWC kinetic rate law precluded its use to determine closed form expressions for the parameters in $v_3$ using either Mathematica or SymPy in Python. In order to overcome this computational obstacle, we modeled the reaction rate for $v_3$ using the convenience kinetic rate law formulation (reference for convenience kinetics). The corresponding expression obtained for $v_3$ is given below (Equation \ref{eq:flux3_convkin}). 
\begin{align}\label{eq:flux3_convkin}
	v_3 = V_3^{max}\left(\frac{1}{1 + \frac{K_3^{pep}}{pep}}\right)\left(\frac{\frac{fdp}{K_3^{fdp}}}{1 + \frac{fdp}{K_3^{fdp}}}\right)
\end{align}

Using this expression for identifiability analysis, we found that each of the parameters $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$ have two different close form expressions owing to the presence of a square root term in their solutions. These distinct expressions are denoted by (1) and (2) following the respective parameter names throughout the rest of the document.

We believe this problem of computational tractability to occur in other complex kinetic rate law formulations as well. Example?

\subsection{Variability in the degree of identifiability of each parameter:}
We use the percentage of data combinations that can identify a parameter as a naive measure of the degree of identifiability of that parameter. As an example, if 90\% of the experimental data combinations used for testing can identify a parameter $\theta_i$, then the degree of identifiability of $\theta_i$ is said to be 0.9 or 90\%. On the other hand, if only 50\% of the combinations can identify another parameter $\theta_j$, then $\theta_j$ has a degree of identifiability of 0.5 or 50\%, and $\theta_i$ has a higher degree of identifiability than $\theta_j$. We can use this criteria to distinguish parameters that can be identified by any type and any combination of experiments from parameters that can be identified by only a select type and combination of experiments.
\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure1/noise_no_noise_kotte_ident_jan01}}
	\caption{The number of data sets, made from a combination of 18 different in silico experiments that can practically identify each parameter in fluxes a) $v_1$, b) $v_2$ and c) $v_3$ when there is no noise in the input experimental data. In the presence of additive noise in the flux and concentration data used for identifiability, the number of data sets practically identifying parameters in fluxes d) $v_1$, e) $v_2$ and f) $v_3$ are shown. 25 samples of additive noise $\epsilon ~ N(0, 0.05)$ were used to generate noisy data for figures in the right hand side panel.}\label{fig:figure1}
\end{figure}

In Figure \ref{fig:figure1} we show the number and percentage of combinations that are capable of identifying each parameter in each flux. In the left panels (Figure \ref{fig:figure1}a-c), the data is shown when noise-free data is used for identification. The identifiability information for parameters using data with additive noise is shown on the right panel (Figure \ref{fig:figure1}d-f). The data shown in the right hand side panel was generated using 25 different samples of random additive noise selected from a normal distribution with 0 mean 0.05 standard deviation. The value used for standard deviation reflects a 5\% noise level. As seen in these figures, irrespective of the presence or absence of noise in the data, the number of combinations that can identify each parameter, and concordantly, the degree of identifiability of each parameter varies widely. Next, we look at trends in the degree of identifiability of each parameter in each flux.

Since both fluxes $v_1$ and $v_2$ are represented by the same enzyme kinetic rate law, we compare the number of data combinations that are capable of identifying the different parameters used to model these fluxes. The number of data combinations identifying the maximum reaction rates ($V_i^{max}$) for both $v_1$ (Figure \ref{fig:figure1}a and d) and $v_2$ (Figure \ref{fig:figure1}b and e) is always greater for than the number of data combinations that identify the enzyme binding ($K_i$) constants in the presence as well as absence of noise in the data. Even for parameters in $v_3$ the percentage of data combinations identifying $V_3^{max} (1)$ and $V_3^{max}(2)$ are greater than the percentage of combinations that can identify $K_3^{fdp}(1)$ and $K_3^{fdp}(2)$ (Figure \ref{fig:figure1}c and f). Hence, all maximum reaction rate parameters in the network have a higher degree of identifiability that their corresponding enzyme binding constants.

The difference between the degrees of identifiability of each parameter within a single flux varies widely depending on the reaction flux. For instance, the difference between the number of combinations that can identify the parameters used to model $v_1$ is less than 2\% (Figure \ref{fig:figure1}a and d). In contrast, a high number of combinations (89.5\%) can identify $V_2^{max}$ in $v_2$, while less than 2\% of the combinations are able to identify $K_2^{pep}$ (Figure \ref{fig:figure1}b and e). This variability in the degree of identifiability for various parameters modeling the same flux is also seen for $v_3$ (Figure \ref{fig:figure1}c and f): $V_3^{max} (1)$ and $V_3^{max} (2)$ have an average degree of 75\% in comparison to the 15\% identifiability for both $K_3^{fdp} (1)$ and $K_3^{fdp} (2)$.

Also notable is the difference in the degree of identifiability between the maximum reaction rates of the uptake flux $v_1$ and the intracellular fluxes $v_2$ and $v_3$. Only about 20\% of the 4896 different data combinations can identify either the maximum reaction rate ($V_1^{max}$) or the enzyme turnover rate ($k_1^{cat}$) in $v_1$ (Figure \ref{fig:figure1}a and d), in contrast to the high number of data combinations capable of identifying $V_2^{max}$ (89.5\%) (Figure \ref{fig:figure1}b and e), $V_3^{max} (1)$ or $V_3^{max} (2)$ (>75\%) (Figure \ref{fig:figure1}c and f).

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight, keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure2/k1cat_no_noise_experiments_jan03_ab}}
	\caption{The contribution of different experiments types used in combinations that can practically identify parameter $k_1^{cat}$ in $v_1$. Panels a) and b) represent the position of the experiment in $k = \{1, 2\}$}\label{fig:figure2}
\end{figure}

This informs us that the degree of identifiability and consequently, the type of experiments needed to identify different parameters varies widely depending the position of the flux in the metabolic network with respect to the inputs and the outputs of the network. The dependency can be further elucidated using the closed form solutions obtained for parameters of $v_1$ in Section \ref{sec:example}. The identifiability expressions for both parameters in $v_1$ (Equation \ref{eq:v1_par}) are dependent on changes in the both the uptake flux $v_1$ as well as the acetate concentration $ac$ in both the experiments used for identification. If either of these values cannot be distinguished between the two experiments whose data is used for identification, then the corresponding combination of experiments can be guaranteed to result in practical non-identifiability of the parameters in $v_1$. As the uptake flux for the metabolic network that is modeled to be dependent on the substrate concentration, the only way meaningful changes in $v_1$ can be observed are through experiments involving changes either in the substrate concentration or changes in the enzyme concentration for $v_1$.
In contrast, any experiment that can distinguish between different values of $pep$ and $v_2$ can be used to identify $V_2^{max}$.

Both in the presence as well as absence of noise, the identifiability of parameters in $v_1$ is guaranteed by only about 20\% of the available data combinations (Figure \ref{fig:figure1}a, d).

 This is further illustrated in Figure below where the different experiments in each combination are shown.

>>>>

\section{Results:}
First in Section \ref{sec:closedform}, we discuss the ability to apply our methods to different nonlinear kinetic rate law formulations within the context of the gluconeogenic model. %Then, in Section we discuss the different experimental design strategies that were found through our identifiability analysis to enable kinetic parameter estimation for the each flux in the model using steady state data. Finally, in Section we expand the application of our method to a relatively large (39 reactions) kinetic model of the red blood cell metabolism.
\subsection{Getting closed-form expressions for each flux in Kotte model:}\label{sec:closedform}
As demonstrated for identifying parameters of $v_1$ earlier in Section \ref{sec:example}, the computational effort required to establish identifiability is governed by the nonlinear complexity of the enzyme kinetic rate law used to model a specific flux. Although computer algebra systems are capable of handling complex symbolic calculations, sometimes, the complexity of getting closed form expressions for all kinetic parameters of certain rate law formulations is too much for the computer algebra system to overcome. We encounter this scenario when expression for the the allosterically activated reaction $v_3$ is described by the MWC model for allosteric regulation. We find that the complexity of the MWC kinetic rate law precludes its use to determine closed form expressions for the parameters in $v_3$ using either Mathematica or SymPy in Python. In order to overcome this computational obstacle, we model the reaction rate for $v_3$ using the convenience kinetic rate law formulation (reference for convenience kinetics). The corresponding expression obtained for $v_3$ is given below (Equation \ref{eq:flux3_convkin}). 
\begin{align}\label{eq:flux3_convkin}
	v_3 = V_3^{max}\left(\frac{1}{1 + \frac{K_3^{pep}}{pep}}\right)\left(\frac{\frac{fdp}{K_3^{fdp}}}{1 + \frac{fdp}{K_3^{fdp}}}\right)
\end{align}

Using this expression for identifiability analysis, we find that each of the parameters $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$ have two different close form expressions owing to the presence of a square root term in their solutions. These distinct expressions are denoted by (1) and (2) following the respective parameter names throughout the rest of the document.

We believe this problem of computational tractability will occur in other complex kinetic rate law formulations as well. Example?

\subsection{Maximum reaction rates are more identifiable than enzyme binding constants:}
We use the percentage of data combinations that can identify a parameter as a simple measure of the degree of identifiability of that parameter. As an example, if 90\% of all the experimental data combinations used for testing can identify a parameter $\theta_i$, then the degree of identifiability of $\theta_i$ is said to be 0.9 or 90\%. On the other hand, if only 50\% of the combinations can identify another parameter $\theta_j$, then $\theta_j$ has a degree of identifiability of 0.5 or 50\%, and $\theta_i$ has a higher degree of identifiability than $\theta_j$. We can use this criteria to distinguish parameters that can be identified by any type and any combination of experiments from parameters that can be identified by only a select type and combination of experiments.

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure1/noise_no_noise_kotte_ident_jan01}}
	\caption{The number of data sets, made from a combination of 18 different in silico experiments that can practically identify each parameter in fluxes a) $v_1$, b) $v_2$ and c) $v_3$ when there is no noise in the input experimental data. In the presence of additive noise in the flux and concentration data used for identifiability, the number of data sets practically identifying parameters in fluxes d) $v_1$, e) $v_2$ and f) $v_3$ are shown. 25 samples of additive noise $\epsilon ~ N(0, 0.05)$ were used to generate noisy data for figures in the right hand side panel. The average and standard deviation values shown for identifiability of parameters with data in the presence of noise were calculated on the basis of the number of data combinations that can identify the respective parameter for each of the 25 noisy data samples.}\label{fig:figure1}
\end{figure}

In Figure \ref{fig:figure1} we show the number and percentage of combinations that are capable of identifying each parameter in each flux. In the left panels (Figure \ref{fig:figure1}a-c), the data is shown when noise-free data is used for identification. The identifiability information for parameters using data with additive noise is shown on the right panel (Figure \ref{fig:figure1}d-f). The data shown in the right hand side panel was generated using 25 different samples of random additive noise selected from a normal distribution with 0 mean 0.05 standard deviation. The value used for standard deviation reflects a 5\% noise level. The average and standard deviation values shown for identifiability of parameters with data in the presence of noise were calculated on the basis of the number of data combinations that can identify the respective parameter for each of the 25 noisy data samples. As seen in these figures, irrespective of the presence or absence of noise in the data, the number of combinations that can identify each parameter, and concordantly, the degree of identifiability of each parameter varies widely. Next, we look at some of the trends in the degree of identifiability of each parameter in each flux.

Since both fluxes $v_1$ and $v_2$ are represented by the same enzyme kinetic rate law, we compare the number of data combinations that are capable of identifying the different parameters used to model these fluxes. The number of data combinations identifying the maximum reaction rates ($V_i^{max}$) for both $v_1$ (Figure \ref{fig:figure1}a and d) and $v_2$ (Figure \ref{fig:figure1}b and e) is always greater than the number of data combinations that identify the enzyme binding ($K_i$) constants. This trend is observed for data combinations with and without noise. Even for parameters in $v_3$ the percentage of data combinations identifying $V_3^{max} (1)$ and $V_3^{max}(2)$ are greater than the percentage of combinations that can identify $K_3^{fdp}(1)$ and $K_3^{fdp}(2)$ (Figure \ref{fig:figure1}c and f). Hence, all maximum reaction rate parameters in the network have a higher degree of identifiability that their corresponding enzyme binding constants.

The difference between the degrees of identifiability of each parameter within a single flux varies widely depending on the reaction flux. For instance, the difference between the percentage of combinations that can identify the parameters used to model $v_1$ is less than 2\% (Figure \ref{fig:figure1}a and d). In contrast, a high number of combinations (89.5\%) can identify $V_2^{max}$ in $v_2$, while less than 2\% of the combinations are able to identify $K_2^{pep}$ (Figure \ref{fig:figure1}b and e). This variability in the degree of identifiability for various parameters modeling the same flux is also seen for $v_3$ (Figure \ref{fig:figure1}c and f): $V_3^{max} (1)$ and $V_3^{max} (2)$ have an average degree of 75\% in comparison to the 15\% identifiability for both $K_3^{fdp} (1)$ and $K_3^{fdp} (2)$. Again, this observation is common for both data with and without added noise.

Also notable is the difference in the degree of identifiability between the maximum reaction rates of the uptake flux $v_1$ and the intracellular fluxes $v_2$ and $v_3$. We discuss this difference and its implications for experimental design in the following section.

\subsection{Experiments required for identifying parameters depends on the position of the flux in the metabolic network:}
Only about 20\% of the 4896 different data combinations can identify either the maximum reaction rate ($V_1^{max}$) or the enzyme turnover rate ($k_1^{cat}$) in $v_1$ (Figure \ref{fig:figure1}a and d), in contrast to the high number of data combinations capable of identifying $V_2^{max}$ (89.5\%) (Figure \ref{fig:figure1}b and e), $V_3^{max} (1)$ or $V_3^{max} (2)$ ($>$75\%) (Figure \ref{fig:figure1}c and f).

This informs us that the degree of identifiability and consequently, the type of experiments needed to identify different parameters varies widely depending the position of the flux in the metabolic network with respect to the inputs and the outputs of the network. The dependency can be further elucidated using the closed form solutions obtained for parameters of $v_1$ in Section \ref{sec:example}. The identifiability expressions for both parameters in $v_1$ ($V_1^{max}$ and $K_1^{ac}$ in Equation \ref{eq:v1_par}) are dependent on changes in the both the uptake flux $v_1$ as well as the acetate concentration $ac$ in both the experiments used for identification. If either of these values cannot be distinguished between the two experiments whose data is used for identification, then the corresponding combination of experiments can be guaranteed to result in practical non-identifiability of the parameters in $v_1$. As the uptake flux for the metabolic network that is modeled to be dependent on the substrate concentration, the only way meaningful changes in $v_1$ can be observed are through experiments involving changes either in the substrate concentration or changes in the enzyme concentration for $v_1$.
In contrast, any experiment that can distinguish between different values of $pep$ and $v_2$ can be used to identify $V_2^{max}$.

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight, keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure2/v1max_k1cat_all_experiments_jan01}}
	\caption{The contribution of different experiments types used in a combination of two experiments ($k = \{1, 2\}$) that can practically identify parameter $V_1^{max}$ a) experiment 1 ($k=1$), and b) experiment 2 ($k=2$) in combination. Different experiments that contribute towards identifiability of $k_1^{cat}$ c) experiment 1 ($k=1$) and d) experiment 2 ($k=2$) in combination.}\label{fig:figure2}
\end{figure}

 This is further illustrated in Figure \ref{fig:figure2} where the different experiments in each combination identifying both $V_1^{max}$ and $k_1^{cat}$ are shown.
